---
title: "PCA plotting of the COL.EUR.AFR.NAT"
author: "Juliana Acosta-Uribe"
date: "5/4/2020"
output:
  html_document:
    df_print: paged
---

# 1. Install and Load Packages

The PCAviz R package provides a simple interface for quickly creating visually compelling plots from Principal Components Analysis (PCA) and accompanying data.

```{r setup, include=FALSE}
#install.packages("devtools")
#devtools::install_github("NovembreLab/PCAviz",build_vignettes = TRUE)

library(PCAviz)
library(ggplot2) # makes pretty plots
library(cowplot)
```

#2. Load the data and the PCA results

Load eigenvectors

```{r echo=TRUE}
Prefix <- "PCAviz"
nPCs <- 10 # NUmber of PCs that were calculated

PCA <-read.table(paste(Prefix, ".eigs", sep = "")) #read table

names(PCA) <- c("ID", paste("PC", (1:nPCs), sep = ""), "case.control") #Add a Header

PCA <- PCA [, 1:(nPCs + 1)] #remove Case.control column
```

Load and edit eigenvalues

```{r echo=TRUE}
eig.val <- sqrt(unlist(read.table(paste(Prefix, ".eval", sep = ""))) [1:nPCs]) # eig.val = square root of eigenvalues

sum.eig <- sum(unlist(read.table(paste(Prefix, ".eval", sep = "")))) #add all eigevalues

```

Load and edit snpweight matrix

```{r echo=TRUE}
snpeigs <-read.table(paste(Prefix, ".snpeigs", sep = ""))

names(snpeigs) <-c("ID", "chr", "pos", paste("PC", (1:nPCs), sep ="")) #Add a Header

snpeigs$chr <- factor(snpeigs$chr)

rownames (snpeigs) <- snpeigs$ID

snpeigs <- snpeigs[, -1]
```

Edit the individual IDs in the PCA file

```{r echo=TRUE}
tmp <- unlist(sapply(as.character(PCA$ID), strsplit, ":"))

ids <- tmp[seq(2, length(tmp), by = 2)]

PCA$ID <- ids
```

Assign a Goup or Cluster to each Individual

```{r echo=TRUE}
ancestry <- read.table("ancestry_proportions.txt", header =TRUE)

ancestry_unord <- ancestry$Group[match(ids, ancestry$ID)]

PCA <- as.data.frame(PCA)

PCA <-cbind(PCA, ancestry_unord)

names(PCA)[ncol(PCA)] <- "sample" 
```

Add the ancestry proportions of each individual (these were calculated with RFMix)
You can add a column to later modify the color that each popultion will have
```{r echo=TRUE}

EUR <- ancestry$EUR[match(ids, ancestry$ID)]
AFR <- ancestry$AFR[match(ids, ancestry$ID)]
NAT <- ancestry$NAT[match(ids, ancestry$ID)]
Ancestry <- ancestry$Ancestry[match(ids, ancestry$ID)]
colors <- ancestry$color[match(ids, ancestry$ID)]

PCA <-cbind(PCA, EUR, AFR, NAT, Ancestry, colors)
```



# 3. Build the PCAviz object

```{r echo=TRUE}

col <- pcaviz(dat = PCA,
              sdev = eig.val,
              var = sum.eig,
              rotation = snpeigs)

col <- pcaviz_abbreviate_var(col, "sample")

```


#4. Graph the PCA

Edit the style for  graphs
```{r echo=TRUE}
geom.point.summary.params <-list(shape = 16, 
                                 stroke = 1,
                                 size = 5,
                                 alpha = 1,
                                 show.legend = FALSE)
#give custom colors to each sample

clrs <-c(with(col$data,tapply(as.character(colors),sample,function (x) x[1])))
names(clrs) <- NULL
```

Plot the PCA!!

```{r echo=TRUE}

plot (col, 
      label="sample",
      colors= clrs ,
      abbreviated.label = FALSE ,
      geom.text.params = list(size = 3,fontface = "plain",na.rm = TRUE),
      geom.point.summary.params = list(shape = 16,stroke = 1,size = 10,
                                       show.legend = FALSE,alpha = 0.6),
      geom.text.summary.params = list(size = 3.25,color = "black",
                                      show.legend = FALSE,alpha = 0.8))
```


You can arrange multiple plots as a grid

```{r echo=TRUE}

plot1 <- plot (col, 
              label="sample",
              abbreviated.label = FALSE,
              colors= clrs,
              coords = paste0("PC", c(1,2)),
              geom.point.summary.params = geom.point.summary.params,
              scale.pc.axes = 0.6)

plot2 <- plot(col,
              label="sample",
              abbreviated.label = FALSE,
              colors= clrs,
              coords = paste0("PC", c(1,3)),
              geom.point.summary.params = geom.point.summary.params,
              scale.pc.axes = 0.6)

plot3 <- plot(col,
              label="sample",
              abbreviated.label = FALSE,
              colors= clrs,
              coords = paste0("PC", c(2,3)),
              geom.point.summary.params = geom.point.summary.params,
              scale.pc.axes = 0.6)

plot4 <- plot(col,
              label="sample",
              abbreviated.label = FALSE,
              colors= clrs,
              coords = paste0("PC", c(3,4)),
              geom.point.summary.params = geom.point.summary.params,
              scale.pc.axes = 0.6)

plot_grid(plot1, plot2, plot3, plot4)
```

As an alternative visualization, it can be helpful to see the distribution of PC coordinates per population for each labeled group in the data

```{r echo=TRUE}

pcaviz_violin(col,
              pc.dims = paste0("PC", c(1:3)),
              plot.grid.params = list(nrow=3))
```

it is useful to inspect the PC loadings to ensure that they broadly represent variation across the genome, rather than one or a small number of genomic regions

```{r echo=TRUE}
for (i in 1:5) {
  plotname <- paste ("plot", i, sep = "")
  plot <-pcaviz_loadingsplot(col,
                             pc.dim = paste0("PC", i),
                             min.rank = 0.8, gap = 200, color = "chr",
                             geom.point.params = list(show.legend = FALSE)) +
                            xlab("SNPs") + ylab (paste0("PC", i, " loading"))
                            assign(plotname, plot)
}

plot <- pcaviz_loadingsplot(col,
                             pc.dim = paste0("PC", 1),
                             min.rank = 0.8, gap = 200, color = "chr") +
                            guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
                            theme(legend.position = "bottom",
                                  legend.justification = "center")
                            
plot_legend <- get_legend(plot)

prow <- plot_grid(plot1, plot2, plot3, plot4, plot5,
                  nrow = 5,
                  align = "vh")

plot_grid(prow, plot_legend, ncol = 1, rel_heights = c(1, .2))

```

Create an Scree plot to represent the variation explained by each PC

```{r echo=TRUE}
screeplot(col, type = "pve") +
  ylim(0, 0.12) +
  ylab("Proportion of Variance Explained") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(axis.line = element_line(size = 1, linetype = "solid"))
```


#5. Analyze PCA according to ancestry percentages


plot_eur <- plot(col, 
                 draw.points = TRUE, 
                 color = "EUR", 
                 group = NULL,
                 shape = "Ancestry")

plot_afr <- plot(col, 
                 draw.points = TRUE, 
                 color= "AFR", 
                 group = NULL,
                 shape = "Ancestry")

plot_nat <- plot(col, 
                 draw.points = TRUE, 
                 color= "NAT", 
                 group = NULL,
                 shape = "Ancestry")

plot_grid(plot_nat,plot_eur,plot_afr,labels = c("A", "B", "C"), nrow=1)



Analyze correlation of PC1 and PC2 with each specific Ancestry



# Plot European Ancestry
plotEUR.PC1 <- plot(col,
                    abbreviated.label = FALSE,
                    coords = c("EUR","PC1"),
                    colors= clrs,
                    group = NULL,
                    show.legend = FALSE)

plotEUR.PC2 <- plot(col,
                    abbreviated.label = FALSE,
                    coords = c("EUR","PC2"),
                    colors= clrs,
                    group = NULL,
                    show.legend = FALSE)

plotEUR <-plot_grid(plotEUR.PC1 , plotEUR.PC2, labels = c("A","B"))


# Plot African Ancestry
plotAFR.PC1 <- plot(col,
                    abbreviated.label = FALSE,
                    coords = c("AFR","PC1"),
                    colors= clrs,
                    group = NULL,
                    show.legend = FALSE)

plotAFR.PC2 <- plot(col,
                    abbreviated.label = FALSE,
                    coords = c("AFR","PC2"),
                    colors= clrs,
                    group = NULL,
                    show.legend = FALSE)

plotAFR <-plot_grid(plotAFR.PC1 , plotAFR.PC2, labels = c("A","B"))


# Plot Native American Ancestry
plotNAT.PC1 <- plot(col,
                    abbreviated.label = FALSE,
                    coords = c("NAT","PC1"),
                    colors= clrs,
                    group = NULL,
                    show.legend = FALSE)

plotNAT.PC2 <- plot(col,
                    abbreviated.label = FALSE,
                    coords = c("NAT","PC2"),
                    colors= clrs,
                    group = NULL,
                    show.legend = FALSE)

plotNAT <-plot_grid(plotNAT.PC1 , plotNAT.PC2, labels = c("A","B"))


plot_grid(plotEUR, plotAFR, plotNAT, ncol = 1)


